Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-21 text/discussion.qmd
tweaking presentation

Second, technological advances created means of identifying and serving prospects. College Board began selling the contact information of test-takers in 1972 [@belkin2019-studata]. ACT followed suit and the two testing companies enjoyed a near duopoly for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another source of student list data is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based -- as opposed to list-based -- targeting. However, key to both approaches is the extent to which the product is built around a large, proprietary database of prospects. This can be observed in the justification for the purchase of Intersect from a University of Utah procurement document [@utah_procurement]:

@@

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company, failing miserably, and being acquired by Nexus Capital.

EAB is the exemplar private equity darling. START HERE ON MONDAY. THE SLIDE IS

The EAB Story: Acquire and Integrate
Acquistiions in lead generation (Jaquette, Salazar, and Martin 2022; Marcus 2024)
Consulting firms sell software-as-service “solutions”
Build subscription recruiting platforms around database of prospects
In their analysis of quantifying school quality in England, @RN5005 [p. 517] observe that "one problem with school league tables \ldots is that the measures of school quality often merely reflect the social origins of those who attend a particular school." Similarly, considering prior research showing that SAT scores are substantially a function of social origin [@RN1273], the Market Segment Model argues that student demand for higher education is mostly a function of social origin.

Institutional theory defines the organizational field as "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services and products" [@RN527, p. 143].

>>>>

Second, technological advances created means of identifying and serving prospects. College Board began selling the contact information of test-takers in 1972 [@belkin2019-studata]. ACT followed suit and the two testing companies enjoyed a near duopoly for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another source of student list data is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based -- as opposed to list-based -- targeting.

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company, failing miserably, and being acquired by Nexus Capital. [ADD A SENTENCE ABOUT COLLEGE BOARD]

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent enrollment management (EM) consulting firm, serving more than 2,100 colleges and offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB are traced to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

As an owner-operated EM consultancy, Royall was the market leader in purchasing student lists on behalf of clients to help realize enrollment goals. Under Vista, EAB tranformed the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours, which can be conceptualized as an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading provider of free college search engine, which is an engine for leads. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista Eqity Partners -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. At the same time, EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform.

EAB combined all of these acquisitions into Enroll360:

> We spent the last couple of years creating a connected recruitment ecosystem that allows enrollment leaders to keep pace with students as they pursue these increasingly digital journeys to college. This work led us to join forces with several leading companies: Cappex, Intersect, Wisr, and YouVisit\ldots Individually, each solution can solve important challenges at various stages of the enrollment funnel\ldots By bringing these capabilities together, our vision is to reinvent how enrollment leaders reach their goals [@koppenheffer_2021]

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50). Thus, the goal of lists was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board *Student Seardh Service* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, Investigative reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. A document from the University of Utah to procure an Intersect subscription explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:

@@

These dynamics of vertical and horizontal acquisitions in the enrollment management market yield policy concerns:

- START HERE TUESDAY: higher prices; funneling prospects only to colleges that are willing/able to pay for subscriptions.

to become a new kind of student list vendor.
